Simultaneous segmentation and distortion correction on diffusion weighted MR using shape priors

### 2012-12-12 Publications/2013-IPMI/Latex/00-Main.tex
Another little step on writing (fast-writing, btw)
Draft update (#25) Latex was not compiling due to incompatibility between the hyperref package and the Springer template. The error is as follows
::
Package hyperref Warning: bookmark level for unknown title defaults to 0.
Package hyperref Warning: bookmark level for unknown author defaults to 0.
Latex Error: ./D1.3.tex:30 Missing \endcsname inserted.

To fix it, just install llncs2e >=2.15
Reviewed equations on methods section. Close #32
Initial draft for the conference. Close #25

\documentclass{llncs}
\usepackage{makeidx}  % allows for indexgeneration
\usepackage{acronym}
\usepackage[numbers]{natbib}
\bibliographystyle{splncsnat}
%\bibliographystyle{IEEEtran}
\usepackage[hyphens]{url}
\usepackage{nameref}
%\rewewcommand\bibname{References}
%

\begin{document}
%
%\tableofcontents
%
\mainmatter              % start of the contributions
%
\title{Simultaneous segmentation and distortion correction on diffusion weighted MR using structural priors}
%
\titlerunning{Segmentation & EPI correction on DWI}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Oscar Esteban\inst{1,2} \and Alessandro Daducci \inst{2} \and
        Virginia Estellers \inst{2} \and Luminita Vese \inst{3} \and 
        Jean-Philippe Thiran \inst{2} \and Andr\'es Santos \inst{1} \and
        Dominique Zosso \inst{2,3}}
%
\authorrunning{Oscar Esteban et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%

>>>>

\documentclass{llncs}
\usepackage{makeidx}  % allows for indexgeneration
\usepackage{acronym}
\usepackage[numbers]{natbib}
\bibliographystyle{splncs03} 
%\bibliographystyle{splncsnat} 
%\bibliographystyle{IEEEtran}

\usepackage[hyphens]{url}
\usepackage{nameref}
\usepackage[breaklinks]{hyperref}
\usepackage[USenglish]{babel}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{pslatex}
%

\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\argmax}{argmax}
\DeclareMathOperator{\argmin}{argmin}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\dist}{dist}
\DeclareMathOperator{\const}{Const.}

\usepackage{graphicx}

%
%\tableofcontents
%
\mainmatter              % start of the contributions
%
\graphicspath{{../Figures/}}

\begin{document}

\title{Simultaneous segmentation and distortion correction on diffusion weighted MR using shape priors}
%%
\titlerunning{Segmentation & EPI correction on DWI}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Oscar Esteban\inst{1,2} \and Alessandro Daducci \inst{2} \and
		Meritxell Bach-Cuadra \inst{2} \and Virginia Estellers \inst{2}
		\and Mar\'ia-Jes\'us Ledesma-Carbayo \and Luminita Vese \inst{3} \and 
        Jean-Philippe Thiran \inst{2} \and Andr\'es Santos \inst{1} \and
        Dominique Zosso \inst{2,3}}
%
\authorrunning{Oscar Esteban et al.} % abbreviated author list (for running head)

### 2012-12-12 Publications/2013-IPMI/Latex/02-Abstract.tex
Initial draft for the conference. Close #25

\begin{abstract}
The abstract should summarize the contents of the paper
using at least 70 and at most 150 words. It will be set in 9-point
font size and be inset 1.0 cm from the right and left margins.
There will be two blank lines before and after the Abstract.
%\dots
%\keywords{diffusion weighted imaging, connectomics, echo planar imaging, magnetic resonance, segmentation, registration, distortion correction}
\end{abstract}

>>>>

\begin{abstract}
Connectomics is the study of fiber connections in the brain, and in particular variations thereof. High-quality reconstructions of the connectome from diffusion weighted MR acquisitions require a precise delineation of the pial surface and the grey-matter-white-matter interface. Also, to enable significant group studies, these surfaces must be located and parcellated consistently across subjects. Precise segmentation and parcellation are readily available on anatomical images obtained e.g., through T1-weighted MR imaging. Diffusion weighted images, however, typically suffer from drastically lower resolution and more important image distortions. Here we propose a joint segmentation-registration framework by exploiting the detailed and consistent anatomy extracted from anatomical MRI as strong shape-prior for DW segmentation. We formulate an instance of Mumford-Shah-like active contours without edges, where we look for a regular deformation field that optimally maps the shape prior on the multivariate features in diffusion space.
%\dots
\keywords{diffusion weighted imaging, connectomics, echo planar imaging, magnetic resonance, segmentation, registration, distortion correction}
\end{abstract}

### 2012-12-12 Publications/2013-IPMI/Latex/03-Introduction.tex
Initial draft for the conference. Close #25

\section{Introduction}
\label{sec:introduction}
%
\ac{dwi} is a widely used family of \ac{mr} techniques
\citep{sundgren_diffusion_2004} which recently has accounted for a growing
interest in its application to whole-brain structural connectivity analysis.
This emerging field, coined in 2005 as \emph{\ac{mr} Connectomics}
\citep{hagmann_diffusion_2005,sporns_human_2005}, currently includes a
large amount of imaging techniques for acquisition, processing and analysis
specifically tuned for the \ac{dwi} data \citep{hagmann_mr_2012}.\\

The whole-brain connectivity analysis has arisen some challenges
that should be overcome in order to get reliable structural information
about the neuronal tracts from \ac{dwi}. The earlier stages of this
processing tools generally include two necessary steps, brain tissue
segmentation on the diffusion space and the correction of geometrical
distortions produced by the imaging techniques.\\

In this work, we will refer as brain tissue segmentation to the precise
delineation of the \ac{csf}-\ac{gm} and \ac{gm}-\ac{wm} interface surfaces.
This segmentation is an important step on which strongly rely further
tasks. In tractography, a high-standard \ac{wm} mask is required. Otherwise,
there is an important risk for the algorithm to lose fiber bundles. This
requirement is usually satisfied by plainly thresholding the \ac{fa}, a
well-know scalar map derived from \ac{dwi} which depicts the isotropy of
water diffusion inside the brain. Additionally, a precise location of the
\ac{gm}-\ac{wm} surface is required in the final steps to
achieve a consistent parcellisation of the cortex to represent the nodes 
of the output network. This parcellisation is generally defined in a 
high-resolution and better understood structural \ac{mri} of the same 
subject (eg. T1 and/or T2 weighted acquisitions). Conversely, this 
problem is resolved with non-linear registration of a structural \ac{mri}
of the subject to the \ac{dwi} data.\\

On the other hand, the \ac{dwi} data is usually obtained with \ac{epi}
acquisition techniques, that often suffer from severe distortions due to 
local field inhomogeneities. Generally, it is easily appreciated in the anterior
part of the brain, along the phase-encoded direction. Some methodologies have
been developed [CITATIONS] and named as \emph{\ac{epi}-unwarp} techniques,
and they require the extra acquisition of the magnitude and phase of
the field (field-mapping), condition which is not always met. Some other 
methodologies do not make use the field-mapping, compensating the distortion
with non-linear registration from structural \ac{mri}.\\

In this paper we propose a novel registration framework to simultaneously
solve the segmentation and distortion challenges, by exploiting as strong 
shape-prior the detailed anatomy extracted from anatomical \ac{mri}. 
We reformulate the segmentation problem as an inverse problem, where
we seek for an underlying deformation field (the distortion) mapping 
from the structural space into the diffusion space.

>>>>

\section{Introduction}
\label{sec:introduction}
%
\ac{dwi} is a widely used family of \ac{mr} techniques
\citep{sundgren_diffusion_2004} which recently has accounted for a growing
interest in its application to whole-brain structural connectivity analysis.
This emerging field, coined in 2005 as \emph{\ac{mr} Connectomics}
\citep{hagmann_diffusion_2005,sporns_human_2005}, currently includes a
large amount of imaging techniques for acquisition, processing, and analysis
specifically tuned for \ac{dwi} data.\\

The whole-brain connectivity analysis has arisen some challenges
that should be overcome in order to get reliable structural information
about the neuronal tracts from \ac{dwi} \cite{johansen-berg_using_2009,
jones_white_2012}. The earlier stages of these
processing pipelines generally include two necessary steps, brain tissue
segmentation on the diffusion space and the correction of geometrical
distortions inherent to the imaging techniques \citep{hagmann_mr_2012}.\\

In this work, we will refer as brain tissue segmentation to the precise
delineation of the \ac{csf}-\ac{gm} and \ac{gm}-\ac{wm} interface surfaces.
This segmentation is an important step on which strongly rely further
tasks. In tractography, a high-standard \ac{wm} mask is required. Otherwise,
there is an important risk for the algorithm to lose fiber bundles. This
requirement is usually solved in practice by plainly thresholding the 
\ac{fa} (a well-know scalar map derived from \ac{dwi} which depicts 
the isotropy of water diffusion inside the brain). 
Additionally, a precise location of the
\ac{gm}-\ac{wm} surface is required in the final steps to
achieve a consistent parcellisation of the cortex to represent the nodes 
of the output network. This parcellisation is generally defined in a 
high-resolution and better understood structural \ac{mri} of the same 
subject (e.g. T1 and/or T2 weighted acquisitions). Conversely, this 
problem is resolved with non-linear registration of a structural \ac{mri}
of the subject to the \ac{dwi} data. Even though some efforts have addressed
the study of the robustness of tractography versus the intra-subject variability
\cite{wakana_reproducibility_2007,heiervang_between_2006}, the results produced
are restricted to relevant regions of the brain. Therefore, extremely robust
and precise segmentation methods are required in the whole-brain application. \\

On the other hand, the \ac{dwi} data is usually obtained with \ac{epi}
acquisition techniques, that often suffer from severe distortions due to 
local field inhomogeneities. Generally, it is easily appreciated in the anterior
part of the brain, along the phase-encoded direction. Some methodologies have
been developed and generically named as \emph{\ac{epi}-unwarp} techniques
\cite{holland_efficient_2010,hsu_correction_2009,jezzard_characterization_2005,
reber_correction_2005}. They usually 
require the extra acquisition of the magnitude and phase of
the field (field-mapping), condition which is not always met. Some other 
methodologies do not make use the field-mapping, compensating the distortion
with non-linear registration from structural \ac{mri} or other means
\citep{andersson_modeling_2001}. To our knowledge, there exists no study
of the impact of the \ac{epi} distortion on the variability of tractography
results. \\

In this paper we propose a novel registration framework to simultaneously
solve the segmentation and distortion challenges, by exploiting as strong 
shape-prior the detailed anatomy extracted from anatomical \ac{mri}. This
is justified by the strong relationship between both problems, and the
advantage of a significant increase of coherence among these steps in
the complete processing pipeline.
We reformulate the segmentation problem as an inverse problem, where
we seek for an underlying deformation field (the distortion) mapping 
from the structural space into the diffusion space.

### 2012-12-12 Publications/2013-IPMI/Latex/04-Methods.tex
Initial draft for the conference. Close #25

\section{Methods}
\label{sec:methods}
%
\subsection{Simulated datasets}
%
As suggested in \autoref{sec:introduction}, the general situation 
consists of having reliable segmentations on the \ac{t1} reference
space, obtained with 
\emph{FreeSurfer\footnote{\url{http://surfer.nmr.mgh.harvard.edu/}}}.
Therefore, regarding the proposed solution, we will have a precise
location of the tissue interfaces of interest in a reference space.
On the other hand, we have a \ac{dwi} volume, characterized by its
low resolution (typically around $2.2x2.2x3mm^3$). Depending on the
posterior reconstruction methodology and the angular resolution
intended, the \ac{dwi} raw data has to be processed in order to
extract the information in a manageable manner. Particularly, we
will use the \ac{fa} and \ac{md} maps. Whereas \ac{fa} describes
the \emph{shape} of diffusion, the \ac{md} depicts the
\emph{manitude} of the process. There exist to main reasons to 
justify their choice. First, they are well-understood and
standardized in clinical routine. Second, they are statistically 
orthogonal and together contain most of the information that is
usually extracted from the \ac{dwi}-derived scalar maps. \\
In order that demonstrating the functionality of the proposed
methodology and characterize its possibilities, we developed two
synthetic model, generating the \ac{dwi} data as described in
 \citep{tuch_q-ball_2004}. We selected 30 directions, for being
a very common protocol for \ac{dti} reconstruction. The first model
is a set of spherical shapes representing the different brain
tissues. The second model is based on the BrainWeb dataset. We
reconstructed the \ac{dwi} data with standard to approximate the
environment to the real one at maximum. There is no interest on the
anatomical reference, given that with the models we hold \emph{a priori}
precisely located surface of the interfaces of interest. \\

FIGURE OF THE MODELS AND EXTRACTED FA, MD

@@

At each iteration, we update the distortion along the steepest 
energy descent. This gradient descent step can be efficiently 
tackled by discretizing the time in a forward Euler scheme, 
and making the right hand side semi-implicit in the 
regularization terms:
%
\begin{align}
\frac{u^{t+1}-u^t}{\tau} &= - \sum_{i=1}^{N_c} \left[ \Delta E(f(c_i'))  \hat{n}_{c_i'} \Psi_{c_i}(x) \right] -\alpha u^{t+1} + \beta\Delta u^{t+1}
\end{align}
%
where the data terms remain functions of the current estimate 
$u^t$, thus $c_i' = c_i'(u^t)$. For simplicity on notation, we 
restricted the number of priors to only 1. We also defined 
$\Delta E(f(c_i')) = E_{out}(f(c_i')) - E_{in}(f(c_i'))$, 
and $E_R(f(c_i')) = (f(c_i')-\mu_R)^T\Sigma_R^{-1}(f(c_i')-\mu_R)$.
We applied a spectral approach to solve this implicit scheme:
%
\begin{equation}
u^{t+1} = \mathcal{F}^{-1}\left\{ \frac{\mathcal{F}\{u^t/\tau
- \sum_{i=1}^{N_c} \left[ \Delta E(f(c_i')) \hat{n}_{c_i'} \Psi_{c_i}(x) \right]  \}}{\mathcal{F}\{(1/\tau+\alpha)\mathcal{I}-\beta\Delta\}} \right\}
\end{equation}
%

\subsection{Experiment}
%
For both models, we created manually a sound distortion visually similar
to real \ac{epi} distortions. We interpolated the distortion to a 
dense deformation field, necessary for warping the raw \ac{dwi} simulated
data. Once the signal was deformed, we proceeded to reconstruct the
\ac{dti} and subsequently obtained the scalars of interest (\ac{fa}, \ac{md}).\\

We evaluate the performance of our methodology to estimate the deformation
field, obtaining a precise segmentation on the diffusion space.

>>>>

\section{Methods}
\label{sec:methods}
%
\subsection{Simulated datasets}
%
As described in \autoref{sec:introduction}, the general situation in
the connectivity pipelines consists on having 
a reliable segmentation obtained from the high resolution \ac{t1} 
reference image. Therefore, a precise location of the tissue interfaces
of interest is available in a reference space. Given that there is no 
interest on the anatomical reference segmentation,
we directly obtained the shape priors from the models. \\

On the other hand, the target \ac{dwi} data is characterized by its
low resolution (typically around $2.2x2.2x3mm^3$). Depending on the
posterior reconstruction methodology and the angular resolution
intended, the \ac{dwi} raw data has to be processed in order to
extract the information in a manageable manner. Particularly, we
will use the \ac{fa} and \ac{md} maps for convenience.
Whereas \ac{fa} describes the \emph{shape} of diffusion, 
the \ac{md} depicts the \emph{manitude} of the process. 
There exist two main reasons to justify their choice. 
First, they are well-understood and standardized in clinical routine.
Second, together they contain most of the information that is
usually extracted from the \ac{dwi}-derived scalar maps. \\

In order that demonstrating the functionality of the proposed
methodology and characterize its possibilities, we developed two
synthetic models and simulated their corresponding \ac{dwi}
raw signal as described in \citep{tuch_q-ball_2004}. 
(HERE WE NEED A GOOD DESCRIPTION OF THE DATA, directions, resolution, etc)
. The first model consists of several spherical shapes emulating
the different brain tissues (see \autoref{fig:fa}, first row). 
The second model is based on the BrainWeb dataset.
We reconstructed the \ac{dwi} signal with standard procedures to approximate the
environment to the real one at maximum. \\

@@

At each iteration, we update the distortion along the steepest 
energy descent. This gradient descent step can be efficiently 
tackled by discretizing the time in a forward Euler scheme, 
and making the right hand side semi-implicit in the 
regularization terms:
%
\begin{align}
\frac{u^{t+1}-u^t}{\tau} &= - \sum_{i=1}^{N_c} \left[ \Delta E(f(c_i'))  \hat{n}_{c_i'} \Psi_{c_i}(x) \right] -\alpha u^{t+1} + \beta\Delta u^{t+1}
\end{align}
%
where the data terms remain functions of the current estimate 
$u^t$, thus $c_i' = c_i'(u^t)$. For simplicity on notation, we 
restricted the number of priors to only 1. We also defined 
$\Delta E(f(c_i')) = E_{out}(f(c_i')) - E_{in}(f(c_i'))$, 
and $E_R(f) = \sqrt{(f-\mu_R)^T\Sigma_R^{-1}(f-\mu_R)}$.
We applied a spectral approach to solve this implicit scheme:
%
\begin{equation}
u^{t+1} = \mathcal{F}^{-1}\left\{ \frac{\mathcal{F}\{u^t/\tau
- \sum_{i=1}^{N_c} \left[ \Delta E(f(c_i')) \hat{n}_{c_i'} \Psi_{c_i}(x) \right]  \}}{\mathcal{F}\{(1/\tau+\alpha)\mathcal{I}-\beta\Delta\}} \right\}
\end{equation}
%

\subsection{Experiment}
%
For both models, we created manually a sound distortion visually similar
to real \ac{epi} distortions. We interpolated the distortion to a 
dense deformation field, necessary for warping the raw \ac{dwi} simulated
data. Once the signal was deformed, we proceeded to reconstruct the
\ac{dti} and subsequently obtained the scalars of interest (\ac{fa}, \ac{md}) and estimated their parameters on the model.\\

We evaluate the performance of our methodology to estimate the deformation
field and compare it to the original synthetic deformation field applied to the data.

### 2012-12-12 Publications/2013-IPMI/Latex/05-Results.tex
Initial draft for the conference. Close #25

\section{Results and discussion}
\label{sec:results}
%

>>>>

\section{Results and discussion}
\label{sec:results}
%
The proposed method successfully reverted the synthetic distortion
field we applied to the data. With 16x16x16 control points, the
displacements field is dense enough to correctly represent the
synthetic field. (INCLUDE FIGURE). Figure XX shows the restored
image and a difference map with the original model (we can also
compute Dice indexes and that stuff).\\
%

\begin{figure}
\begin{tabular}{ccccc}
\includegraphics[width=0.19\textwidth]{fig_modelFA_00} &
\includegraphics[width=0.19\textwidth]{fig_modelFA_01} &
\includegraphics[width=0.19\textwidth]{fig_modelFA_02} &
\includegraphics[width=0.19\textwidth]{fig_modelFA_03} &
\includegraphics[width=0.19\textwidth]{fig_modelFA_04} \\
\includegraphics[width=0.19\textwidth]{fig_contours_00} &
\includegraphics[width=0.19\textwidth]{fig_contours_01} &
\includegraphics[width=0.19\textwidth]{fig_contours_02} &
\includegraphics[width=0.19\textwidth]{fig_contours_03} &
\includegraphics[width=0.19\textwidth]{fig_contours_04}
\end{tabular}
\caption{First row presents several slices along Z axis of the \ac{fa} map obtained 
after \ac{dti} fitting (using the original signal without simulated distortion) and
the corresponding \ac{wm}-\ac{gm} contour. Second row contains the warped \ac{fa}
map (using the distorted signal). The red contour corresponds to the interface
detected after registration. In blue color, the initial \ac{wm}-\ac{gm} contour
(the same as in previous row).}
\label{fig:fa}
\end{figure}
